---
title: "Exploratory Analysis of Spaceship Titanic"
author: "Monica Amezquita"
date: "2022-07-29"
output: pdf_document
---

## Load Relevant Libraries and Data

```{r echo=FALSE}
library(stringr)
library(corrplot)
library(GGally)
library(plotly)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# set working directory to project directory (outside of SRC folder)
split <- str_split(getwd(), "/", simplify = TRUE)
wdir <- paste(split[-length(split)], collapse = "/")
knitr::opts_knit$set(root.dir = wdir)
```


# Get Training Data with New Features

```{r}
# get training set
train <- read.csv(sprintf("%s/data/imputed_feature_train.csv", getwd()), na.strings = "")
```

# Explore the Training Set

```{r}
#ggpairs(train[-c(5,8,16,17)])  # ignore features with many levels (i.e. names)
```

# Percentage of Transported

```{r}
proportions <- table(train$Transported)/length(train$Transported)
percentages <- proportions*100

```

# Explore Expenditure vs. Transported and Expenditure vs. Crysleep colored by Transported

```{r}
ggplot(train, aes(x = Expenditure,fill = Transported)) + 
  geom_histogram() + 
  facet_wrap(vars(Transported))


ggplot(train[!train$CryoSleep,], aes(x = CryoSleep,y = Expenditure, fill = Transported)) + 
  geom_boxplot() + 
  ylim(c(0,4000))
```

# Explore Count of Individuals in each Deck colored by Transported

```{r}
ggplot(train, aes(x = DeckNum, fill = Transported)) + 
  geom_bar() + 
  facet_wrap(vars(Transported))
```


# Explore Count of "Side" colored by Transported
```{r}
ggplot(train, aes(x = DeckNum, fill = Transported)) + 
  geom_bar() + 
  facet_wrap(vars(Transported))
```

# Explore Histogran of Age colored by IsSoloTraveler
```{r}
ggplot(train, aes(x = Age, fill = IsSoloTraveler)) + 
  geom_histogram(binwidth = 1) + 
  facet_wrap(vars(Transported))
```

# Explore Histogram of Age colored by Transported
```{r}
ggplot(train, aes(x = Age, fill = Transported)) + 
  geom_histogram() +
  facet_wrap(vars(Transported))
```

```{r}
ggplot(train, aes(x = Transported, fill = IsSoloTraveler)) + 
  geom_histogram() +
  facet_wrap(vars(IsSoloTraveler))
```

```{r}
ggplot(train, aes(x = IsSoloChild, fill = Transported)) + 
  geom_histogram() +
  facet_wrap(vars(Transported))
```

```{r}
ggplot(train, aes(x = Deck, fill = CryoSleep)) + 
  geom_bar() + 
  facet_wrap(vars(Transported)) + 
  facet_wrap(vars(CryoSleep))


ggplot(train[!train$CryoSleep,], aes(x = CryoSleep,y = Expenditure, fill = Transported)) + 
  geom_boxplot() + 
  ylim(c(0,4000))
```

```{r}
transported_bycabin_cryosleep <- sqldf("Select deck, count(deck) as count_deck from train where transported = 1 and cryosleep = true group by deck")
```
