---
title: "Imputation with Spaceship Titanic"
author: "Alejandro Hernandez"
date: "2022-08-08"
output: pdf_document
---
```{r, include=FALSE}
rm(list=ls()) # clear all local variables
library(stringr)  # string manipulation
library(dplyr)  # data frame manipulation
library(mice) # imputation functions
library(ggplot2)  # plotting
library(tidyr)  # replacing NA

# trash me later pls
library(caret)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# set working directory to project directory (outside of SRC folder)
split <- str_split(getwd(), "/", simplify = TRUE)
wdir <- paste(split[-length(split)], collapse = "/")
knitr::opts_knit$set(root.dir = wdir)
```

*Imputation* is the practice of assigning a missing value to an observation/record by inference or prediction. The goal of this script is to impute the necessary fields of the *Spaceship Titanic* dataset. Preferred methods of imputation vary by dataset, so we aim to utilize multiple methods.

Below is a complete implementation of our imputation pipeline.

# Full Pipeline

First, let's define functions we use later:

```{r}
# to read in the data
get_data <- function(df_name){
  # get training set
  data <- read.csv(sprintf("%s/data/%s.csv", getwd(), df_name), na.strings = "")

  # split aggregate columns
  data <- separate(data, col=Cabin, into=c("Deck", "Num", "Side"), sep='/', fill="right")
  data <-  separate(data, col=PassengerId , into=c("GroupId", "PersonalId"), sep='_', fill="right")
  data <-  separate(data, col=Name , into=c("FirstName", "LastName"), sep=' ', fill="right")
  
  data[data == "NA"] <- NA
  
  # convert logical characters to type LOGICAL
  vars <- c('CryoSleep', 'VIP')
  data[vars] <- sapply(data[vars], as.integer)
  
  # convert numeric characters to type INTEGER
  vars <- c('GroupId', 'PersonalId', 'Age', 'RoomService', 'FoodCourt', 'ShoppingMall', 'Spa', 'VRDeck')
  data[vars] <- sapply(data[vars], as.integer)
  
  # convert missing first and last names to "NONE"
  # imputation treats NA, and it wouldn't make sense to impute
  data <- data %>%
    mutate_at(c('FirstName','LastName'), ~ replace_na(.,"NONE"))
  
  # encode columns of type LOGICAL to INTEGER (TRUE = 1, FALSE = 0)
  data <- data %>%
    mutate_if(is.logical, as.integer)
  return(data)
}
```

```{r}
# to get mode of a list
get_mode <- function(v, na.rm = FALSE) {
  v <- ifelse(anyNA(v), na.omit(v), v)
  uniqv <- unique(v)
  mode <- uniqv[which.max(tabulate(match(v, uniqv)))]
  return(mode)
}
```

Now, we can execute our complete imputation pipeline.
First, we need to create our 3 data sets.

```{r}
# ---- initialize training/validation/testing data sets ----
set.seed(28765)
original_train <- read.csv(sprintf("%s/data/train.csv", getwd()), na.strings = "")
train_index <- createDataPartition(original_train$Transported, 
                                   p = .75)$Resample1
train <- original_train[train_index, ]
valid <- original_train[-train_index,]
write.csv(train,
          sprintf("%s/data/training.csv", getwd()), row.names = FALSE)
write.csv(valid,
          sprintf("%s/data/validation.csv", getwd()), row.names = FALSE)

train <- get_data(df_name = "training")
valid <- get_data(df_name = "validation")
test <- get_data(df_name = "test")
```

### Group-based Imputation

From exploratory analysis, we found that the values of some predictors (e.g. **HomePlanet**, **Deck**, **Destination**, **ShoppingMall**, etc.) are similar among members of a group.

It would be reasonable then, to consider a group's most common planet of departure, destination, etc. as the best estimate of that group's member.

Below is our implementation of this method. Note that if a group is a single passengers, whose value for a selected variable is NA, that observation cannot be estimated.

```{r}
# to impute validation and testing data from training data
group_imputation <- function(train, valid, test){
  # impute HomePlanet by group
  na_groups <- unique(train$GroupId[is.na(train$HomePlanet)])
  df <- train$GroupId[train$GroupId %in% na_groups]
  na_groups <- names(table(df)[table(df) > 1])
  for (groupid in na_groups) {
    mode <- get_mode(train$HomePlanet[train$GroupId == groupid])
    train$HomePlanet <- ifelse(train$GroupId == groupid & is.na(train$HomePlanet), mode, train$HomePlanet)
    valid$HomePlanet <- ifelse(valid$GroupId == groupid & is.na(valid$HomePlanet), mode, valid$HomePlanet)
    test$HomePlanet <- ifelse(test$GroupId == groupid & is.na(test$HomePlanet), mode, test$HomePlanet)
  }
  # impute Deck by group
  na_groups <- unique(train$GroupId[is.na(train$Deck)])
  df <- train$GroupId[train$GroupId %in% na_groups]
  na_groups <- names(table(df)[table(df) > 1])
  for (groupid in na_groups) {
    mode <- get_mode(train$Deck[train$GroupId == groupid])
    train$Deck <- ifelse(train$GroupId == groupid & is.na(train$Deck), mode, train$Deck)
    valid$Deck <- ifelse(valid$GroupId == groupid & is.na(valid$Deck), mode, valid$Deck)
    test$Deck <- ifelse(test$GroupId == groupid & is.na(test$Deck), mode, test$Deck)
  }
  # impute Side by group
  na_groups <- unique(train$GroupId[is.na(train$Side)])
  df <- train$GroupId[train$GroupId %in% na_groups]
  na_groups <- names(table(df)[table(df) > 1])
  for (groupid in na_groups) {
    mode <- get_mode(train$Side[train$GroupId == groupid])
    train$Side <- ifelse(train$GroupId == groupid & is.na(train$Side), mode, train$Side)
    valid$Side <- ifelse(valid$GroupId == groupid & is.na(valid$Side), mode, valid$Side)
    test$Side <- ifelse(test$GroupId == groupid & is.na(test$Side), mode, test$Side)
  }
  # impute Destination by group
  na_groups <- unique(train$GroupId[is.na(train$Destination)])
  df <- train$GroupId[train$GroupId %in% na_groups]
  na_groups <- names(table(df)[table(df) > 1])
  for (groupid in na_groups) {
    mode <- get_mode(train$Destination[train$GroupId == groupid])
    train$Destination <- ifelse(train$GroupId == groupid & is.na(train$Destination), mode, train$Destination)
    valid$Destination <- ifelse(valid$GroupId == groupid & is.na(valid$Destination), mode, valid$Destination)
    test$Destination <- ifelse(test$GroupId == groupid & is.na(test$Destination), mode, test$Destination)
  }
  # impute VIP by group
  na_groups <- unique(train$GroupId[is.na(train$VIP)])
  df <- train$GroupId[train$GroupId %in% na_groups]
  na_groups <- names(table(df)[table(df) > 1])
  for (groupid in na_groups) {
    mode <- get_mode(train$VIP[train$GroupId == groupid])
    train$VIP <- ifelse(train$GroupId == groupid & is.na(train$VIP), mode, train$VIP)
    valid$VIP <- ifelse(valid$GroupId == groupid & is.na(valid$VIP), mode, valid$VIP)
    test$VIP <- ifelse(test$GroupId == groupid & is.na(test$VIP), mode, test$VIP)
  }
  # impute CryoSleep by group
  na_groups <- unique(train$GroupId[is.na(train$CryoSleep)])
  df <- train$GroupId[train$GroupId %in% na_groups]
  na_groups <- names(table(df)[table(df) > 1])
  for (groupid in na_groups) {
    mode <- get_mode(train$CryoSleep[train$GroupId == groupid])
    train$CryoSleep <- ifelse(train$GroupId == groupid & is.na(train$CryoSleep), mode, train$CryoSleep)
    valid$CryoSleep <- ifelse(valid$GroupId == groupid & is.na(valid$CryoSleep), mode, valid$CryoSleep)
    test$CryoSleep <- ifelse(test$GroupId == groupid & is.na(test$CryoSleep), mode, test$CryoSleep)
  }
  # impute RoomService by group
  na_groups <- unique(train$GroupId[is.na(train$RoomService)])
  df <- train$GroupId[train$GroupId %in% na_groups]
  na_groups <- names(table(df)[table(df) > 1])
  for (groupid in na_groups) {
    avg <- mean(train$RoomService[train$GroupId == groupid], na.rm = TRUE)
    train$RoomService <- ifelse(train$GroupId == groupid & is.na(train$RoomService), avg, train$RoomService)
    valid$RoomService <- ifelse(valid$GroupId == groupid & is.na(valid$RoomService), avg, valid$RoomService)
    test$RoomService <- ifelse(test$GroupId == groupid & is.na(test$RoomService), avg, test$RoomService)
  }
  # impute ShoppingMall by group
  na_groups <- unique(train$GroupId[is.na(train$ShoppingMall)])
  df <- train$GroupId[train$GroupId %in% na_groups]
  na_groups <- names(table(df)[table(df) > 1])
  for (groupid in na_groups) {
    avg <- mean(train$ShoppingMall[train$GroupId == groupid], na.rm = TRUE)
    train$ShoppingMall <- ifelse(train$GroupId == groupid & is.na(train$ShoppingMall), avg, train$ShoppingMall)
    valid$ShoppingMall <- ifelse(valid$GroupId == groupid & is.na(valid$ShoppingMall), avg, valid$ShoppingMall)
    test$ShoppingMall <- ifelse(test$GroupId == groupid & is.na(test$ShoppingMall), avg, test$ShoppingMall)
  }
  return(list('train'=train, 'valid'=valid, 'test'=test))
}
```

```{r}
group_imputations <- group_imputation(train, valid, test)
train_imp <- group_imputations$train
valid_imp <- group_imputations$valid
test_imp <- group_imputations$test
```

### Planet-based Imputation

Now, any passenger with missing data for **HomePlanet** and **Destination** is traveling alone. To impute their values, we can identify the most common destination for each home planet (across all passengers), and vice versa.

```{r}
# get levels of home and destination planet
homes <- levels(factor(train_imp$HomePlanet))
h_modes <- c()
dests <- levels(factor(train_imp$Destination))
d_modes <- c()
# for each home/destination, compute the mode destination/home
for (i in 1:length(homes)) {
  h_mode <- get_mode(c(train_imp$Destination[train_imp$HomePlanet == homes[i]]))
  h_modes <- append(h_modes, h_mode)
  
  d_mode <- get_mode(c(train_imp$HomePlanet[train_imp$Destination == dests[i]]))
  d_modes <- append(d_modes, d_mode)
}
home_df <- data.frame("Home" = homes, "ModeDestination"  = h_modes)
dest_df <- data.frame("Destination" = dests, "ModeHomePlanet"  = d_modes)

# TRAINING DATA
# fill in value of HomePlanet based on mode Destination
for (dest in dests) {
  train_imp$HomePlanet <- ifelse(
    is.na(train_imp$HomePlanet) & train_imp$Destination == dest, 
    dest_df$ModeHomePlanet[dest_df$Destination == dest],
    train_imp$HomePlanet)
}
# fill in value of Destination based on mode HomePlanet
for (home in homes) {
  train_imp$Destination <- ifelse(
    is.na(train_imp$Destination) & train_imp$HomePlanet == home, 
    home_df$ModeDestination[home_df$Home == home],
    train_imp$Destination)
}

# VALDIATION DATA
# fill in value of HomePlanet based on mode Destination
for (dest in dests) {
  valid_imp$HomePlanet <- ifelse(
    is.na(valid_imp$HomePlanet) & valid_imp$Destination == dest, 
    dest_df$ModeHomePlanet[dest_df$Destination == dest],
    valid_imp$HomePlanet)
}
# fill in value of Destination based on mode HomePlanet
for (home in homes) {
  valid_imp$Destination <- ifelse(
    is.na(valid_imp$Destination) & valid_imp$HomePlanet == home, 
    home_df$ModeDestination[home_df$Home == home],
    valid_imp$Destination)
}

# TESTING DATA
# fill in value of HomePlanet based on mode Destination
for (dest in dests) {
  test_imp$HomePlanet <- ifelse(
    is.na(test_imp$HomePlanet) & test_imp$Destination == dest, 
    dest_df$ModeHomePlanet[dest_df$Destination == dest],
    test_imp$HomePlanet)
}
# fill in value of Destination based on mode HomePlanet
for (home in homes) {
  test_imp$Destination <- ifelse(
    is.na(test_imp$Destination) & test_imp$HomePlanet == home, 
    home_df$ModeDestination[home_df$Home == home],
    test_imp$Destination)
}
```

## Multivariate Imputation by Chained Equations (MICE)

Whichever records have not been completely imputed, impute with multidimensional parametric models with the MICE library.
After, save each imputed data set to a CSV.

The **mice** package allows us to make strong imputation of numeric, binary, and ordinal variables.
It provides access to many powerful models. In our case, we selected the folowing models for the following data types:

* Numeric: multivariate linear regression + lasso penalty
* Binary: multivariate logistic regression + lasso penalty
* Ordinal: linear discriminant analysis (LDA)

This selection was informed by our familiarity with the models and our experience with their success are estimators.

```{r}
# train and validation data
train_valid_df <- rbind(train_imp, valid_imp)
# convert categorical variables (to be imputed) to factors
vars <- c("HomePlanet", "Deck", "Side", "Destination", "VIP", "CryoSleep")
for (var in vars) {
  train_valid_df[,var] <- factor(train_valid_df[,var])
}
train_valid_df$Num <- as.integer(train_valid_df$Num)
# impute using column-dependent methods
mice_mod <- mice(train_valid_df[!names(train_valid_df) %in% c('Transported')], 
                 method=c("", "", # GroupId, PersonalId
                          "lda", # HomePlanet
                          "lasso.logreg", # CryoSleep
                          "lda", # Deck
                          "lasso.norm", # Num
                          "lasso.logreg", # Side
                          "lda", # Destination
                          "lasso.norm", # Age
                          "lasso.logreg", # VIP
                          "lasso.norm", # Room Service
                          "lasso.norm", # Food Court
                          "lasso.norm", # Shopping Mall
                          "lasso.norm", # Spa
                          "lasso.norm", # VRDeck
                          "", "" # First Name, Last Name
                          ),
                 ignore = c(rep(FALSE, nrow(train)), rep(TRUE, nrow(valid))))
train_valid_imp <- complete(mice_mod)
train_valid_imp[c('Transported')] <- train_valid_df[c('Transported')]
train_imp <- train_valid_imp[1:nrow(train),]
valid_imp <- train_valid_imp[-c(1:nrow(train)),]
# the numeric variables (such as Age, RoomService, etc.) were predicted as negative
# we'll adjust all negative elements in the data frame to be zero
valid_imp[valid_imp < 0] <- 0
train_imp[train_imp < 0] <- 0
write.csv(train_imp, sprintf("%s/data/train_imputed.csv", getwd()), row.names = FALSE)
write.csv(valid_imp, sprintf("%s/data/valid_imputed.csv", getwd()), row.names = FALSE)

# test data
test_imp$CryoSleep <- as.logical(test_imp$CryoSleep)
test_imp$VIP <- as.logical(test_imp$VIP)
test_imp[c('Transported')] <- rep(-1, nrow(test_imp))
train_test_df <- rbind(train_imp, test_imp)
# convert categorical variables (to be imputed) to factors
vars <- c("HomePlanet", "Deck", "Side", "Destination", "VIP", "CryoSleep")
for (var in vars) {
  train_test_df[,var] <- factor(train_test_df[,var])
}
train_test_df$Num <- as.integer(train_test_df$Num)
# impute using column-dependent methods
mice_mod <- mice(train_test_df[!names(train_test_df) %in% c('Transported')], 
                 method=c("", "", # GroupId, PersonalId
                          "lda", # HomePlanet
                          "lasso.logreg", # CryoSleep
                          "lda", # Deck
                          "lasso.norm", # Num
                          "lasso.logreg", # Side
                          "lda", # Destination
                          "lasso.norm", # Age
                          "lasso.logreg", # VIP
                          "lasso.norm", # Room Service
                          "lasso.norm", # Food Court
                          "lasso.norm", # Shopping Mall
                          "lasso.norm", # Spa
                          "lasso.norm", # VRDeck
                          "", ""
                          ),
                 ignore = c(rep(FALSE, nrow(train)), rep(TRUE, nrow(test))))
train_test_imp <- complete(mice_mod)
train_test_imp['Transported'] <- train_test_df$Transported
test_imp <- train_test_imp[-c(1:nrow(train)),]
# the numeric variables (such as Age, RoomService, etc.) were predicted as negative
# we'll adjust all negative elements in the data frame to be zero
test_imp[test_imp < 0] <- 0
test_imp$Transported <- train_test_imp$Transported[-c(1:nrow(train))]

write.csv(test_imp, sprintf("%s/data/test_imputed.csv", getwd()), row.names = FALSE)
```

All done! To confirm all NAs have been estimated, count before/after NAs.

```{r}
# view number of NA in each variable pre-imputation
colSums(is.na(train))
cat("\n\n")
# view number of NA in each variable post-imputation
colSums(is.na(train_imp))
```

```{r}
# view number of NA in each variable pre-imputation
colSums(is.na(valid))
cat("\n\n")
# view number of NA in each variable post-imputation
colSums(is.na(valid_imp))
```

```{r}
# view number of NA in each variable pre-imputation
colSums(is.na(test))
cat("\n\n")
# view number of NA in each variable post-imputation
colSums(is.na(test_imp))
```

# Supplementray

## Mean/Mode Imputation

Filling missing values of a variable with the variable's mean (if numeric) or mode (if categorical) is a quick, simple, and easy strategy to implement that seems- at a glance- reasonable. However, there are some serious drawbacks to this method. For numeric data, because mean imputed values are totally independent from other predictors, they have the potential to erase correlation between variables.

```{r}
# define function to compute mode
getmode <- function(v, na.rm = FALSE) {
  v <- ifelse(anyNA(v), na.omit(v), v)
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}
```

```{r}
# fill mean for numeric columns
mean_imp <- train %>%
  mutate_all(~ifelse(is.na(.x),
                     mean(.x, na.rm = TRUE),
                     .x))
# fill mode for categorical columns
mean_imp <- mean_imp %>%
  mutate_all(~ifelse(is.na(.x),
                     getmode(.x, na.rm = TRUE),
                     .x))

# confirm all NA are gone
anyNA(mean_imp)
```

## Confirming Group Similarity

Below is code used to select which expediture variables to estimate from group averages.

```{r}
# FOR NA Groups
# compute average stand.dev and var of RoomService by groups
na_groups <- unique(group_imp$GroupId[is.na(group_imp$RoomService)])
df <- group_imp$GroupId[group_imp$GroupId %in% na_groups]
na_groups <- names(table(df)[table(df) > 1])
var_arr <- c()
for (groupid in na_groups) {
  ls <- c(group_imp$RoomService[group_imp$GroupId == groupid])
  var_arr <- append(var_arr, var(ls, na.rm = TRUE))
}
sd_arr <- sqrt(var_arr)
cat("RoomService\nMean var:", mean(var_arr, na.rm = TRUE), "\nMean stand dev:", mean(sd_arr, na.rm = TRUE))

# compute average stand.dev and var of FoodCourt by groups
na_groups <- unique(group_imp$GroupId[is.na(group_imp$FoodCourt)])
df <- group_imp$GroupId[group_imp$GroupId %in% na_groups]
na_groups <- names(table(df)[table(df) > 1])
var_arr <- c()
for (groupid in na_groups) {
  ls <- c(group_imp$FoodCourt[group_imp$GroupId == groupid])
  var_arr <- append(var_arr, var(ls, na.rm = TRUE))
}
sd_arr <- sqrt(var_arr)
cat("\n\nFoodCourt\nMean var:", mean(var_arr, na.rm = TRUE), "\nMean stand dev:", mean(sd_arr, na.rm = TRUE))

# compute average stand.dev and var of ShoppingMall by groups
na_groups <- unique(group_imp$GroupId[is.na(group_imp$ShoppingMall)])
df <- group_imp$GroupId[group_imp$GroupId %in% na_groups]
na_groups <- names(table(df)[table(df) > 1])
var_arr <- c()
for (groupid in na_groups) {
  ls <- c(group_imp$ShoppingMall[group_imp$GroupId == groupid])
  var_arr <- append(var_arr, var(ls, na.rm = TRUE))
}
sd_arr <- sqrt(var_arr)
cat("\n\nShoppingMall\nMean var:", mean(var_arr, na.rm = TRUE), "\nMean stand dev:", mean(sd_arr, na.rm = TRUE))

# compute average stand.dev and var of Spa by groups
na_groups <- unique(group_imp$GroupId[is.na(group_imp$Spa)])
df <- group_imp$GroupId[group_imp$GroupId %in% na_groups]
na_groups <- names(table(df)[table(df) > 1])
var_arr <- c()
for (groupid in na_groups) {
  ls <- c(group_imp$Spa[group_imp$GroupId == groupid])
  var_arr <- append(var_arr, var(ls, na.rm = TRUE))
}
sd_arr <- sqrt(var_arr)
cat("\n\nSpa\nMean var:", mean(var_arr, na.rm = TRUE), "\nMean stand dev:", mean(sd_arr, na.rm = TRUE))

# compute average stand.dev and var of VRDeck by groups
na_groups <- unique(group_imp$GroupId[is.na(group_imp$VRDeck)])
df <- group_imp$GroupId[group_imp$GroupId %in% na_groups]
na_groups <- names(table(df)[table(df) > 1])
var_arr <- c()
for (groupid in na_groups) {
  ls <- c(group_imp$VRDeck[group_imp$GroupId == groupid])
  var_arr <- append(var_arr, var(ls, na.rm = TRUE))
}
sd_arr <- sqrt(var_arr)
cat("\n\nVRDeck\nMean var:", mean(var_arr, na.rm = TRUE), "\nMean stand dev:", mean(sd_arr, na.rm = TRUE))
```

**End of Imputation.**
