---
title: "Model Building"
author: "Monica Amezquita; Alejandro Hernandez; Hugo Marquez"
date: "2022-08-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# set working directory to project directory (outside of SRC folder)
split <- str_split(getwd(), "/", simplify = TRUE)
wdir <- paste(split[-length(split)], collapse = "/")
knitr::opts_knit$set(root.dir = wdir)
```

```{r setup, include=FALSE}
library(stringr)
library(mice)
library(dplyr)
library(caret)
library(yardstick)
```

# Get Training Data with New Features

```{r}
# get training set
train <- read.csv(sprintf("%s/data/imputed_feature_train.csv", getwd()), na.strings = "")
validation <- read.csv(sprintf("%s/data/imputed_feature_validation.csv", getwd()),na.strings = "")
test <- read.csv(sprintf("%s/data/imputed_feature_test.csv", getwd()), na.strings = "") 
```

# Model Building

# Data Pre-Processing

### Split training and validation set

```{r}

space_titanic_train <- train
space_titanic_valid <- validation
space_titanic_test <- test

cat("Training set size: ", nrow(space_titanic_train))
cat("\nValidation set size: ", nrow(space_titanic_valid))
```

### Binary Classification

```{r}

#Adding levels for Training Dataset
space_titanic_train$Transported <- factor(space_titanic_train$Transported)
levels(space_titanic_train$Transported) <- c("No", "Yes")


#Adding levels for Validation Dataset
space_titanic_valid$Transported <- factor(space_titanic_valid$Transported)
levels(space_titanic_valid$Transported) <- c("No", "Yes")
```

## K-Nearest Neighbors

```{r}
control <- trainControl(method="cv",number=10)

tr_control <- trainControl(method="cv",number=10, classProbs = TRUE)


fit_knn <- caret::train(Transported ~  HomePlanet + CryoSleep + DeckNum  + Side + Destination + Age + Expenditure + people_in_group,
                 method = "knn",
                 trControl = control,
                 preProcess = c("center","scale"),
                 data = space_titanic_train)

# add predictions to training data frame
space_titanic_train <- space_titanic_train %>%
  add_predictions(fit_knn, var = "knn_class") %>%
  mutate(knn_prob = predict(fit_knn, newdata = ., type="prob")$Yes)

# add predictions to validation data frame
space_titanic_valid <- space_titanic_valid %>%
  add_predictions(fit_knn, var = "knn_class") %>%
  mutate(knn_prob = predict(fit_knn, newdata = ., type="prob")$Yes)

(fit_knn$results)
```

```{r}
## Sensitivity and Specificity Metrics
metrics <- metric_set(accuracy, sens, spec, ppv, npv)

space_titanic_valid %>% 
  metrics(estimate = Transported, truth = knn_class)
```

## Confusion Matrix

```{r}
space_titanic_valid %>% 
  conf_mat(Transported, knn_class)  
```

## ROC Curve and AUC

```{r}
# Train ROC Plot 
space_titanic_train %>% 
  roc_curve(truth = Transported,
            estimate = knn_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4) + 
  ggtitle("ROC Plot for Training Dataset Using KNN")

# Test ROC Plot 
space_titanic_valid %>% 
  roc_curve(truth = Transported,
            estimate = knn_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4) + 
  ggtitle("ROC Plot for Validation Dataset Using KNN")
```

## Least Squares Estimate 

```{r}
fit_lr <- caret::train(Transported ~  HomePlanet + CryoSleep + DeckNum  + Side + Destination + Age + Expenditure + people_in_group,
                method = "glm",
                trControl = control,
                data = space_titanic_train)

# add predictions to training data frame
space_titanic_train <- space_titanic_train %>%
  add_predictions(fit_lr, var = "lr_class") %>%
  mutate(lr_prob = predict(fit_lr, newdata = ., type="prob")$Yes)

# add predictions to validation data frame
space_titanic_valid <- space_titanic_valid %>%
  add_predictions(fit_lr, var = "lr_class") %>%
  mutate(lr_prob = predict(fit_lr, newdata = ., type="prob")$Yes)

(fit_lr$results)

coef(fit_lr$finalModel)
```

### Sensitivity and Specificity Metrics

```{r}
metrics <- metric_set(accuracy, sens, spec, ppv, npv)

space_titanic_valid %>% 
  metrics(estimate = Transported, truth = lr_class)
```

### Confusion Matrix

```{r}
space_titanic_valid %>% 
  conf_mat(Transported, lr_class)  
```

### ROC Curve and AUC

```{r}
#Train ROC 
space_titanic_train %>% 
  roc_curve(truth = Transported,
            estimate = lr_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4) + 
  ggtitle("ROC Plot for Training Dataset Using Least Squares")

#Test ROC 
space_titanic_valid %>% 
  roc_curve(truth = Transported,
            estimate = lr_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4) + 
  ggtitle("ROC Plot for Validation Dataset Using Least Squares")
```


## Lasso and Ridge Classification

```{r}

tr_control <- trainControl(method="cv", number = 5, classProbs=TRUE)

lasso <- caret::train(Transported ~ HomePlanet + CryoSleep + DeckNum  + Side + Destination + Age + Expenditure + people_in_group,
             data = space_titanic_train,
             method = 'glmnet', 
             tuneGrid = expand.grid(alpha = 1, lambda = 1),
             trControl = tr_control 
)

lambda_grid = 10^(seq(-3,1,by=0.1))

ridge <- caret::train(Transported ~ HomePlanet + CryoSleep + DeckNum  + Side + Destination + Age + Expenditure + people_in_group,
             data = space_titanic_train,
             method = 'glmnet', 
             tuneGrid = expand.grid(alpha = 0, lambda = lambda_grid),
             trControl = tr_control
)

space_titanic_train <- space_titanic_train %>%
  spread_predictions(lasso,ridge)

space_titanic_valid <- space_titanic_valid %>%
  spread_predictions(lasso,ridge)
```
## Look at Model Coefficients, Predictions, and Metrics

```{r}
coef(lasso$finalModel, lasso$finalModel$lambdaOpt)
```

```{r}
coef(ridge$finalModel, ridge$finalModel$lambdaOpt)
```

```{r}
#Train Metrics
space_titanic_train %>% metrics(truth = Transported, estimate = lasso)

space_titanic_train %>% metrics(truth = Transported, estimate = ridge)

# Validation Metrics
space_titanic_valid %>% metrics(truth = Transported, estimate = lasso)

space_titanic_valid %>% metrics(truth = Transported, estimate = ridge)
```

## Logistic Regression

```{r}
train_log <- caret::train(Transported ~ HomePlanet + CryoSleep + DeckNum  + Side + Destination + Age + Expenditure + people_in_group, 
                   method = "glm",
                   data=space_titanic_train)

# training set
space_titanic_train <- space_titanic_train %>%
  add_predictions(train_log, var="log_pred") %>% # class predictions
  mutate(log_prob = predict(train_log, newdata = ., type="prob")$Yes) 


# test set
space_titanic_valid <- space_titanic_valid %>%
  add_predictions(train_log, var="log_pred") %>% # class predictions
  mutate(log_prob = predict(train_log, newdata = ., type="prob")$Yes) 
```

### Sensitivity and Specificity Metrics

```{r}

metrics <- metric_set(accuracy, sens, spec, ppv, npv)

space_titanic_valid %>% 
  metrics(estimate = Transported, truth = log_pred)
```

### Confusion Matrix

```{r}
space_titanic_valid %>% 
  conf_mat(Transported, log_pred)  
```

### ROC Curve and AUC

```{r}
#Train Dataset
space_titanic_train %>% 
  roc_curve(truth = Transported,
            estimate = log_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4) + 
  ggtitle("ROC Plot for Training Dataset Using Logistic Regression")

#Validation Dataset
space_titanic_valid %>% 
  roc_curve(truth = Transported,
            estimate = log_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4) + 
  ggtitle("ROC Plot for Validation Dataset Using Logistic Regression")
```

## Support Vector Machine

### Linear Decision Boundary

```{r}
train_svm_lin <- caret::train(Transported ~ HomePlanet + CryoSleep + DeckNum  + Side + Destination + Age + Expenditure + people_in_group,
                       method = "svmLinear", 
                       trControl = tr_control,  
                       data = space_titanic_train,
                       preProcess = c("center","scale"),  # data should be preprocessed for SVM
                       tuneGrid = expand.grid(C = seq(0.01, 2, length=20))  # tuning parameters go here
                   )

# training set
space_titanic_train <- space_titanic_train %>%
  add_predictions(train_svm_lin, var="svm_lin_pred") %>% # class predictions
  mutate(svm_lin_prob = predict(train_svm_lin, newdata = ., type="prob")$Yes) 

# test set
space_titanic_valid <- space_titanic_valid %>%
  add_predictions(train_svm_lin, var="svm_lin_pred") %>% # class predictions
  mutate(svm_lin_prob = predict(train_svm_lin, newdata = ., type="prob")$Yes) 
```

### Sensitivity and Specificity Metrics

```{r}
metrics <- metric_set(accuracy, sens, spec, ppv, npv)

space_titanic_valid %>% 
  metrics(estimate = Transported, truth = svm_lin_pred)
```

### Confusion Matrix

```{r}
space_titanic_valid %>% 
  conf_mat(Transported, svm_lin_pred)  
```

### ROC Curve and AUC

```{r}
#Train ROC
space_titanic_train %>% 
  roc_curve(truth = Transported,
            estimate = svm_lin_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4) + 
  ggtitle("ROC Plot for Training Dataset Using Linear SVM")

#Validation ROC
space_titanic_valid %>% 
  roc_curve(truth = Transported,
            estimate = svm_lin_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4) + 
  ggtitle("ROC Plot for Validation Dataset Using Linear SVM")
```

### Polynomial Decision Boundary

```{r}
train_svm_poly <- caret::train(Transported ~ HomePlanet + CryoSleep + DeckNum  + Side + Destination + Age + Expenditure + people_in_group,
                        method = "svmPoly",
                        trControl = tr_control,
                        data = space_titanic_train,
                        preProcess = c("center","scale"),
                        tuneGrid = expand.grid(scale = seq(0.01,2, length=5),
                                                C = seq(0.01, 2, length=5),
                                                degree = 2))

# training set
space_titanic_train <- space_titanic_train %>%
  add_predictions(train_svm_poly, var="svm_poly_pred") %>% # class predictions
  mutate(svm_poly_prob = predict(train_svm_poly, newdata = ., type="prob")$Yes) 

# test set
space_titanic_valid <- space_titanic_valid %>%
  add_predictions(train_svm_poly, var="svm_poly_pred") %>% # class predictions
  mutate(svm_poly_prob = predict(train_svm_poly, newdata = ., type="prob")$Yes) 


plot(train_svm_poly)
```

### Sensitivity and Specificity Metrics

```{r}

metrics <- metric_set(accuracy, sens, spec, ppv, npv)

space_titanic_valid %>% 
  metrics(estimate = Transported, truth = svm_poly_pred)
```

### Confusion Matrix

```{r}
space_titanic_valid %>% 
  conf_mat(Transported, svm_poly_pred)  
```

### ROC Curve and AUC

```{r}

#Train ROC
space_titanic_train %>% 
  roc_curve(truth = Transported,
            estimate = svm_poly_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4) + 
  ggtitle("ROC Plot for Training Dataset Using SVM Poly")

#Validation ROC

space_titanic_valid %>% 
  roc_curve(truth = Transported,
            estimate = svm_poly_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4)+ 
  ggtitle("ROC Plot for Validation Dataset Using SVM Poly")
```


### Amorphous Decision Boundaries (RBF Kernel)

```{r}
train_svm_rbf <- train(Transported ~ HomePlanet + CryoSleep + DeckNum  + Side + Destination + Age + Expenditure + people_in_group,
                        method = "svmRadial",
                        trControl = tr_control,
                        data = space_titanic_train,
                        preProcess = c("center","scale"),
                        tuneGrid = expand.grid(sigma = seq(0.01,2, length=5),
                                                C = seq(0.01, 2, length=5)))
# training set
space_titanic_train <- space_titanic_train %>%
  add_predictions(train_svm_rbf, var="svm_rbf_pred") %>% # class predictions
  mutate(svm_rbf_prob = predict(train_svm_rbf, newdata = ., type="prob")$Yes) 

# test set
space_titanic_valid <- space_titanic_valid %>%
  add_predictions(train_svm_rbf, var="svm_rbf_pred") %>% # class predictions
  mutate(svm_rbf_prob = predict(train_svm_rbf, newdata = ., type="prob")$Yes) 
```


### Sensitivity and Specificity Metrics

```{r}

metrics <- metric_set(accuracy, sens, spec, ppv, npv)

space_titanic_valid %>% 
  metrics(estimate = Transported, truth = svm_rbf_pred)
```

### Confusion Matrix

```{r}
space_titanic_valid %>% 
  conf_mat(Transported, svm_rbf_pred)  
```

### ROC Curve and AUC

```{r}

#Train ROC
space_titanic_train %>% 
  roc_curve(truth = Transported,
            estimate = svm_rbf_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4) + 
  ggtitle("ROC Plot for Training Dataset Using SVM RBF")

#Test ROC
space_titanic_valid %>% 
  roc_curve(truth = Transported,
            estimate = svm_rbf_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4)+ 
  ggtitle("ROC Plot for Validation Dataset Using SVM RBF")
```


## Fully-Connected Neural Network

```{r}
nn15 <- caret::train(Transported ~ GroupId ,
              method = "mlp",
              preProc = c("center", "scale", "nzv"),
              linOut = TRUE, # if you want regression 
              trControl = tr_control,
              learnFunc = "Rprop", # robust propogation
              maxit = 500, # for long it will run for fitting 
              hiddenActFunc = "Act_ReLU",
              tuneGrid = expand.grid(
                size = c(15) # 5 nodes in hidden layer
              ),
              data = space_titanic_train
              )

# training set
space_titanic_train <- space_titanic_train %>%
  add_predictions(nn15, var="nn_pred") %>% # class predictions
  mutate(nn_prob = predict(nn15, newdata = ., type="prob")$Yes) 

# test set
space_titanic_valid <- space_titanic_valid %>%
  add_predictions(nn15, var="nn_pred") %>% # class predictions
  mutate(nn_prob = predict(nn15, newdata = ., type="prob")$Yes) 
```

### Sensitivity and Specificity Metrics

```{r}
metrics <- metric_set(accuracy, sens, spec, ppv, npv)

space_titanic_valid %>% 
  metrics(estimate = Transported, truth = nn_pred)
```

### Confusion Matrix

```{r}
space_titanic_valid %>% 
  conf_mat(Transported, nn_pred)  
```

### ROC Curve and AUC

```{r}
space_titanic_valid %>% 
  roc_curve(truth = Transported,
            estimate = nn_prob,
            event_level = "second") %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, size = 0.4)
```


# ROC Curves

```{r}
space_titanic_train %>%
  pivot_longer(
    cols = c("knn_prob", "log_prob",
             "lr_prob","svm_lin_prob",
             "svm_poly_prob","svm_rbf_prob","nn_prob"),
    names_to = "model",
    values_to = "prob"
  ) %>%
  group_by(model) %>%
  roc_curve(truth = Transported,
            estimate = prob,
            event_level = "second") %>%
  ggplot(aes(
    x = 1 - specificity,
    y = sensitivity,
    color = model
  )) +
  geom_line() +
  geom_abline(slope = 1,
              intercept = 0,
              size = 0.4) +
  ggtitle("Train Set ROC Curves")
```

```{r}
space_titanic_valid %>%
  pivot_longer(
    cols = c("knn_prob", "log_prob",
             "lr_prob","svm_lin_prob",
             "svm_poly_prob","svm_rbf_prob","nn_prob"),
    names_to = "model",
    values_to = "prob"
  ) %>%
  group_by(model) %>%
  roc_curve(truth = Transported,
            estimate = prob,
            event_level = "second") %>%
  ggplot(aes(
    x = 1 - specificity,
    y = sensitivity,
    color = model
  )) +
  geom_line() +
  geom_abline(slope = 1,
              intercept = 0,
              size = 0.4) +i
  ggtitle("Test Set ROC Curves")
```